---
title: "Inflammation and depression"
author: "Gibran Hemani"
date: "2022-02-10"
output: html_notebook
---

```{r, message=FALSE, warning=FALSE}
library(ieugwasr)
library(tidyverse)
library(TwoSampleMR)
library(rsnps)
```


## Summary

- Not much support for CRP on causal pathway to MDD, modest protective effect of higher inflammation
- Either using all variants or just CRP or IL6R variants
- MVMR of CRP + BMI on MDD shows BMI associates independent of CRP
- Reverse MR of MDD on CRP is null, modest positive effect of higher MDD liability with higher CRP

## Analysis

Depression GWAS:

```{r}
gwasinfo("ieu-b-102") %>% str()
```

C-reactive protein GWAS:

```{r}
gwasinfo("ieu-b-35") %>% str()
```

Perform MR

```{r}
dat <- make_dat("ieu-b-35", "ieu-b-102")
res <- mr(dat)
mr_scatter_plot(res, dat)
```

```{r}
res %>% select(method, nsnp, b, se, pval)
```

Use only CRP variant

```{r}
ress <- mr_singlesnp(dat)
mr_forest_plot(ress)
```

Find CRP SNP

```{r}
crp_snp <- subset(dat, chr == 1) %>%
  mutate(dist=abs(pos.exposure - 159682079)) %>%
  arrange(dist) %>%
  select(dist, SNP, pval.exposure) %>%
  {.$SNP[1]}
```

Find IL6R SNP

```{r}
dbsnp_info <- ncbi_snp_query(dat$SNP)
il6r_snp <- subset(dbsnp_info, gene == "IL6R")$rsid
```
MR using just CRP

```{r}
subset(ress, SNP == crp_snp) %>% str()
```
MR using just IL6R

```{r}
subset(ress, SNP == il6r_snp) %>% str()
```

Multivariable MR of CRP and BMI on MDD

```{r}
mvexp <- mv_extract_exposures(c("ieu-b-40", "ieu-b-35"))
mvout <- extract_outcome_data(mvexp$SNP, "ieu-b-102")
mvdat <- mv_harmonise_data(mvexp, mvout)
mvres <- mv_multiple(mvdat) 
mvres$result %>% select(exposure, nsnp, b, se, pval)
```

Reverse MR

```{r}
datr <- make_dat("ieu-b-102", "ieu-b-35")
resr <- mr(datr)
mr_scatter_plot(resr, datr)
```

```{r}
resr %>% select(method, nsnp,b, se, pval)
```

## Human immune system traits from Roederer et al 2015


```{r}
gi <- gwasinfo()
hi_exp <- extract_instruments(gi$id[grepl("met-b-", gi$id)])
str(hi_exp)
```

```{r}
hi_out <- extract_outcome_data(hi_exp$SNP, "ieu-b-102")
hi_dat <- harmonise_data(hi_exp, hi_out)
hi_res <- mr(hi_dat)
```

Does anything surpass 5% FDR threshold?

```{r}
subset(hi_res, method %in% c("Inverse variance weighted", "Wald ratio")) %>% mutate(fdr=p.adjust(pval, "fdr")) %>%
  arrange(fdr) %>%
  filter(fdr < 0.05) %>%
  select(exposure, nsnp, b, se, pval)
```

```{r}

```

