---
title: "Direct vs confounder instruments"
author: "Gibran Hemani"
date: "2022-02-09"
output: html_notebook
---

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(simulateGP)
library(dplyr)
library(TwoSampleMR)
```


Simulate a model where genotypes influence X and U and U has some influence on X. What are the effects of genotypes on X? What fraction of GWAS hits are direct effects for each percentile of magnitude of G-X assoc?

```{r}
nd <- 1000
nc <- 1000
np <- 100

bxy <- 0.5
bgx <- abs(rnorm(nd))
bgu <- abs(rnorm(nc))

bgux <- bgu * bxy

d <- tibble(b = c(bgx, bgux), path=c(rep("direct", nd), rep("confounder", each=nc))) %>%
	arrange(b) %>%
	mutate(ord=1:n(), perc = ntile(b, np)) %>%
	group_by(perc) %>%
	summarise(b=mean(b), n=n(), prop=sum(path=="direct")/n())

d %>%
ggplot(., aes(x=perc, y=b)) +
geom_jitter(aes(colour=prop)) +
scale_colour_gradient(low="red", high="blue") +
labs(y="Genetic effect on X", x="Percentile", colour="Probability of GWAS hit being a direct effect") +
theme(legend.position="bottom")
```

Figure is used here: https://drive.google.com/file/d/1XW9f0cUUt8jGZWmea11Nb3kN2JsR0sQ5/view?usp=sharing

## Bias from G-U instruments

This is the expected effect estimate of X on Y using an instrument that arises via U:

$$
\beta_{IV,u} = \frac{\beta_{gu} \beta_{uy} + \beta_{gu} \beta_{ux} \beta_{xy}}{\beta_{gu} \beta_{ux}}
$$
which simplifies to

$$
\beta_{IV,u} = \frac{\beta_{uy}}{\beta_{ux}} + \beta_{xy}
$$
Check that this is correct

```{r, message=FALSE}
nid <- 10000
gx <- rbinom(nid, 2, 0.3)
gu <- rbinom(nid, 2, 0.3)

u <- gu + rnorm(nid)
xl <- gx + rnorm(nid)

param <- expand.grid(
	bux = seq(-1, 1, by=0.2),
	buy = seq(-1, 1, by=0.2),
	bxy = c(0, 0.5)
)

out <- lapply(1:nrow(param), function(i)
{
	x <- xl + u * param$bux[i]
	y <- x * param$bxy[i] + u * param$buy[i] + rnorm(nid)
	p <- param[i,]
	p$bgx <- get_effs(x, y, matrix(gx, nid, 1)) %>% mr() %>% {.$b}
	p$bgu <- get_effs(x, y, matrix(gu, nid, 1)) %>% mr() %>% {.$b}
	return(p)
}) %>% bind_rows()

out$exp_gu <- out$buy / out$bux + out$bxy
plot(bgu ~ exp_gu, subset(out, !is.infinite(exp_gu)))
```


